pSEOBNRv4HM

### 2020-10-24 papers/intro_paper/intro_paper.tex
updated paper

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave observations~\citep{GWTC-2} from the first half of their third observing run~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford, Washington and Livingston, Louisiana~\citep{aasi2015characterization}, and the Advanced Virgo detector in Cascina, Italy~\citep{acernese2014advanced} have now detected more than $40$ gravitational wave events from the merger of compact objects like neutron stars and/or black holes (compact binary coalescences/CBCs). Alongside independent claims of detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020}, this has firmly established the field of gravitational wave astronomy, five years on from the first ever direct detection of gravitational waves on Earth, on September 14, 2015~\citep{abbott2016observation}.

@@

Observation of gravitational waves have had significant astrophysical and cosmological implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR, a CBC involving two black holes (a binary black hole (BBH) system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  We have performed tests of gravitational wave generation and source dynamics, where we place bounds on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of gravitational wave propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. We have also checked for consistency between different portions of the signal using estimates of final mass and spin~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR. \abhi{Perhaps reframe this section in the contect of the O3a TGR companion paper}

@@

Tests of black hole ringdown and nature of the remnant object is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as the underlying theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. \abhi{It is a spinning multipolar effective-one-body waveform calibrated to numerical relativity simulations (SEOBNRv4HM~\citep{Cotesta:2018fcv}})This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with black holes spins in the current paper. All astrophysical black holes are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties.
\par
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies in a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real gravitational wave events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section~\ref{sec:discussion}.

@@

We initially demonstrate our method using simulated GR signals, with parameters similar to the two gravitational wave events, GW150914 CITE and GW190521 CITE. These two systems are representative data points for the kind of systems this test is most suitable for: high-mass events which are loud enough that the signal-to-noise ratio pre- and post-merger return reliable parameter estimation results. To avoid possible systematic biases in our parameter estimation results due to error in modelling, we use the GR version of the same waveform, SEOBNRv4HM CITE (without allowing for deviations in the QNM parameters) to simulate our GW signal. And to avoid systematic biases due to noise, we use an averaged (zero-noise) realisation of the Advanced LIGO zero-detuned-high-power PSD CITE. We find, as expected, that the posterior distribution on the parameters describing fractional deviations in the frequency and damping time, $\{d\omega,d\tau\}$ are consistent with zero (the solid curves in left panel of Fig.\ref{fig:gr_simulation}). The difference in the size of the posterior is due to the difference in the SNRs of the two events ($\sim 25$ for GW150914 as compared to $\sim 13$ for GW190521). One can then convert these fractional quantities into absolute quantities using the relations:

@@

\begin{figure*}[h]
	\includegraphics[width=0.3\textwidth]{figures/GW150914_GW190521_simulated_frac.png}\includegraphics[width=0.56\textwidth]{figures/GW150914_GW190521_simulated_abs.png}\label{fig:gr_simulation}
	\caption{final figure will contain a third set of contours for modGR recovered assuming GR to show bias. That run is ongoing.}
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{figures/modGR_waveforms.png}\label{fig:nongr_waveform}
	\caption{NonGR waveforms}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.4\textwidth]{figures/O1O2_realevents.png}
	\includegraphics[width=0.4\textwidth]{figures/GW150914.png}
	\caption{O1-O2 analysis}\label{fig:o1o2_events}
\end{figure}

@@

Tests of general relativity

\begin{figure}[H]
	\includegraphics[width=0.4\textwidth]{figures/S190521g_swinjs.png}
	\caption{a single figure for both Gaussian and Real noise injections, color coded, to save space}\label{fig:21g_systematics}
\end{figure}

>>>>

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave (GW) observations~\citep{GWTC-2} from the first half of their third observing run (O3a)~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford, Washington and Livingston, Louisiana~\citep{aasi2015characterization}, and the Advanced Virgo detector in Cascina, Italy~\citep{acernese2014advanced} have now detected more than $40$ GW events from the merger of compact objects like neutron stars and/or black holes (compact binary coalescences or CBCs). Alongside independent claims of detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020}, this has firmly established the field of GW astronomy, five years on from the first ever direct detection of GWs on Earth, on September 14, 2015~\citep{abbott2016observation}.

@@

Observation of GWs have had significant astrophysical and cosmological implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity (GR)~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR, a CBC involving two black holes (a binary black hole (BBH) system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  

@@

The LIGO-Virgo collaborations have also released a companion paper detailing their latest results of tests of GR using events from the catalogue, GWTC-2. The results include tests of GW generation and source dynamics, where bounds are placed on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of GW propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. The paper also checks for consistency between different portions of the signal using estimates of final mass and spin~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR.
The O3a TGR paper also details results on tests of black hole ringdown and nature of the remnant object, which is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as the underlying theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. \abhi{It is a spinning multipolar effective-one-body waveform calibrated to numerical relativity simulations (SEOBNRv4HM~\citep{Cotesta:2018fcv}})This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with black holes spins in the current paper. All astrophysical black holes are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties.
\par
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies in a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real GW events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section~\ref{sec:discussion}.

@@

We initially demonstrate our method using simulated GR signals, with parameters similar to the two GW events, GW150914 CITE and GW190521 CITE. These two systems are representative data points for the kind of systems this test is most suitable for: high-mass events which are loud enough that the signal-to-noise ratio pre- and post-merger return reliable parameter estimation results. To avoid possible systematic biases in our parameter estimation results due to error in modelling, we use the GR version of the same waveform, SEOBNRv4HM CITE (without allowing for deviations in the QNM parameters) to simulate our GW signal. And to avoid systematic biases due to noise, we use an averaged (zero-noise) realisation of the Advanced LIGO zero-detuned-high-power PSD CITE. We find, as expected, that the posterior distribution on the parameters describing fractional deviations in the frequency and damping time, $\{d\omega,d\tau\}$ are consistent with zero (the solid curves in left panel of Fig.\ref{fig:gr_simulation}). The difference in the size of the posterior is due to the difference in the SNRs of the two events ($\sim 25$ for GW150914 as compared to $\sim 13$ for GW190521). One can then convert these fractional quantities into absolute quantities using the relations:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
	\includegraphics[width=0.3\textwidth]{figures/GW150914_GW190521_simulated_frac.png}\includegraphics[width=0.56\textwidth]{figures/GW150914_GW190521_simulated_abs.png}\label{fig:gr_simulation}
	\caption{final figure will contain a third set of contours for modGR recovered assuming GR to show bias. That run is ongoing.}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
	\includegraphics[width=0.5\textwidth]{figures/modGR_waveforms.png}\label{fig:nongr_waveform}
	\caption{NonGR waveforms}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
	\includegraphics[width=0.4\textwidth]{figures/O1O2_realevents.png}
	\includegraphics[width=0.4\textwidth]{figures/GW150914.png}
	\caption{O1-O2 analysis}\label{fig:o1o2_events}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In order to investigate the

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
	\includegraphics[width=0.4\textwidth]{figures/S190521g_swinjs.png}
	\caption{a single figure for both Gaussian and Real noise injections, color coded, to save space}\label{fig:21g_systematics}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
